Missing the Point: Non-Convergence in Iterative Imputation Algorithms

### 2020-04-27 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
After spellcheck

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
- name: H. I. Oberman
  num: 1
address:
- num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multiple imputation by chained equations (MICE) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
  - Royal
  - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
  rticles::sage_article:
    keep_tex: yes
---

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous, but also tedious. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI yields unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].

@@

<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to ... multiple imputation by chained equations ... "The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.") -->

@@

In this paper we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice]. The convergence properties of this MI algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.

@@

<!-- This is repeated $m$ times. Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$, . Each of the $m$ chains starts with an initial value, drawn randomly from $y_{obs}$. The chains are terminated after a predefined number of iterations. , and subsequently used in the analysis and pooling steps.  -->
<!-- The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'.  -->

@@

The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13]. Without mixing, imputation chains do not intermingle nicely. Without stationarity, there is trending within imputation chains, see Figure \ref{fig:non-conv}. **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable. The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model. Each line is an imputation. In the first plot, the lines intermingle nicely and there is little trending. In the second, there is a lot of trending and the lines do not cross much. Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm. This shows the importance of reaching convergend states in iterative imputation algorithms.**

@@

Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p. 147]. In this study we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one-another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. **Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.

## In practice/Answer
**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condtion must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance). And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**

**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together. So are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up in the typical convergence situation. The superior performance of the typical convergence over the pathological non-convergnce is less prominent, even flipped for $t<4$. Panel C displays the $AC$  as computed with the R function stats::acf(). When we look at this panel, we must conclude that something weird is going on. The $AC$-values indicate equal performance (upto $t=5$) for the typical convergence and the pathological non-convergnce, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending upto iteration number seven. After consulting the documentation on stats::acf(), we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to ... the function works when the number of iterations > 50. The default number of iterations in mice, however,  is 5. Therefore, we compute $AC$ manually, see panel D. The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

@@

Convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$). We expect that a completely convergenced state will not be reached.

@@

The two challenges of convergence are stationarity and mixing. Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances). As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). A user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by [@mack03], we will use introduce a $\theta$ that summarizes the multivariate state space of the algorithm. Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.

@@

The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands (quantities of scientific interest; $Q$s). It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased (and confidence valid) estimate of $Q$, the algorithm is sufficiently converged. For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research. The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**). Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$. Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**). **remove the term performance measure here and add an extra alinea to define performance measures as  bias in all estimands, and coverage rate and CI length of regression coefficients.**

**Move this to Results:** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s. Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases. The point at wich an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ. This means that the algorithm has converged sufficiently (**under the current specifications**).

@@

Evaluation with the performance measures shows that are had an autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.

@@

Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to). In practice thresholds are used to diagnose non-convergence. Usually, non-convergence is diagnosed when $\widehat{R}$  > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero. Performance measures are the same as above: unbiased, confidence valid estimates.

**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is not observed. But what about the thresholds for practice? $\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s. Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again. If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed. The threshold $\widehat{R}$ < 1.1 is somewhat conservative with respect to the performance measures. But it seems OK. $\widehat{R}$ < 1.01 is too stringent compared to the performance measures. And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).

@@

```
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
  for (number of simulation runs from 1 to 1000)
    2. create missingness
    for (number of iterations from 1 to 100)
      3. impute missingness
      4. compute convergence diagnostics
      5. perform analysis of scientific interest
      6. pool results across imputations
      7. compute performance measures
  8. aggregate outcomes across simulation runs 
9. combine outcomes of all missingness proportions
```
The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2, ..., 100$), and 2) varying the missingness proportions ($p_{miss} = .05, .25, .50, .75, .95$). The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints. **As the number of iterations goes up, the imputation chains will become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem. The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions then cause slower convergence. However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.

@@

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim  \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} 
For the purpose of this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd]. A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`. Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem. Let 
$$
Y_i =  1 + 2X_{1i} + .5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $i = 1, 2, ..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.

We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely  ... solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are!**):

@@

The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`. (**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). This study only considers only an MCAR missingness mechanism. Therefore, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. This is just a proof of concept.

Missing datapoints are imputed with the function `mice::mice()`. All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`). The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$).

<!-- What to evaluate? -->

<!-- - bias in regression coefficients -->

<!-- - bias in $R^2$ (= coefficient of determination) -->

<!-- -  -->

<!-- From the imputation step, we extract chain means and chain variances across iterations ($\theta$ at $t = 1,2,\dots,T$).   -->
<!-- We compute convergence diagnostics separately for both $\theta$s, and for each of the four variables in the dataset. We report the maximum (absolute) value as the 'worst linear predictor' of convergence.  -->
$\widehat{R}$ (**of what?? i.e., chain means and chain variances of each variable**) is computed by implementing Vehtari et al.'s [-@veht19] recommendations. $AC$ (**of what? and how is $AC$ aggregated across imputations? i.e., mean in latest sim**) is computed with function `stats::acf()`.

To estimate the quantity of scientific interest, $Q$, we perform multiple linear regression on each completed dataset with the function `stats::lm()`. We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$. We use the function `mice::pool()` to estimate u bar (**check definition!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14.

We compute bias as the difference between $Q$ and $\bar{Q}$. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI95\%) around $\bar{Q}$. We compute the CI95\% bounds as

$$\bar{Q} \pm t_{(m-1)} \times SE_{\bar{Q}},$$

where $t_{(m-1)}$ is the quantile of a $t$-distribution with $m-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. **Why track CIW?** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences.
From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

@@

The bias in the estimates of the variable means show little to no difference between simulation conditions. It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased. Similarly, the bias in the estimated variances is more or less stable across simulation conditions. These univariate quantities appear to be unaffected by the number of iterations.

When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions. Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01). The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$. (With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
<!-- $\widehat{R}$ values are indistinguishable between simulation conditions when $T>30$.  -->
<!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations.  -->

Auto-correlations indicate no sign of trending within imputation chains. In most simulation conditions $AC$ is smaller than or about equal to zero. Across simulation conditions, however, the autocorrelation curve does not trend towards 0. Autocorrelation values plateau off at a value of around .1 . This is a small positive autocorrelation, which would indicate some trending within chains.

The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends and are therefore not discussed separately. Taken together,  univariate estimates seem robust across simulation conditions. There is no clear effect of the number of iterations on the bias in these estimates, while the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).

>>>>

---
title: "Missing the Point: Non-Convergence in Iterative Imputation Algorithms"
runninghead: Oberman
author:
  name: H. I. Oberman
  num: 1
address:
  num: 1
  org: Department of Methodology and Statistics, Utrecht University, Utrecht, The Netherlands
corrauth: "Hanne Oberman, Sjoerd Groenman building, Utrecht Science Park, Utrecht, The Netherlands."
email: h.i.oberman@uu.nl
abstract: "(**Rewrite:**) Multiple imputation by chained equations (MICE) is a widely used tool to accommodate missing data. While empirical evidence supports the validity of inferences obtained using mice, there is no consensus on the convergence properties of the method. This paper provides insight into non-convergence of mice algorithms."
keywords: MICE; convergence 
classoption:
 - Royal
 - times
bibliography: thesis.bib
bibliographystyle: sageh
output:
 rticles::sage_article:
  keep_tex: yes
---

@@

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Anyone who analyzes person-data may run into a missing data problem. Missing data is not only ubiquitous but also tedious. If a dataset contains just one incomplete observation, statistical inferences are undefined and will not produce any results. To circumvent this, many statistical packages employ list-wise deletion by default (i.e., ignoring incomplete observations). Unfortunately, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) the missing values in the incomplete observations. Subsequently, statistical inferences can be performed on the completed dataset. By repeating this process several times, a distribution of plausible results may be obtained, which reflects the uncertainty in the data due to missingness. This technique is known as 'multiple imputation' [MI; @rubin76]. MI yields unbiased and confidence valid estimates of the true---but missing---data inference under many circumstances [@buur18].

@@

<!-- (**Rewrite:** With FCS, an iterative algorithmic procedure is used to... multiple imputation by chained equations... "The values for imputing the missing data are drawn from the posterior predictive distribution of the missing data conditional on the observed data.") -->

@@

In this paper, we investigate how non-convergence in iterative imputation algorithms may be diagnosed, how well these methods perform, and at which point convergence may safely be assumed. For reasons of brevity, we only focus on the MI algorithm implemented in the world-leading MI software: the `R` [@R] package `mice` [@mice]. The convergence properties of this MI algorithm are investigated through model-based simulation. The results of this simulation study are guidelines for assessing convergence of MI algorithms, which will aid applied researchers in drawing valid inference from incomplete datasets.

@@

<!-- This is repeated $m$ times. Imputed values are the ultimate samples of a `chain' of For each imputed value ($t = T$), a s for each missing datapoint in $y_{mis}$,. Each of the $m$ chains starts with an initial value, drawn randomly from $y_{obs}$. The chains are terminated after a predefined number of iterations. , and subsequently used in the analysis and pooling steps. -->
<!-- The collection of samples between the initial value (at $t=1$) and the imputed value (at $t=T$) will be referred to as an 'imputation chain'. -->

@@

The two challenges for convergence of iterative algorithms are mixing and stationarity [@gelm13]. Without mixing, imputation chains do not intermingle nicely. Without stationarity, there is trending within imputation chains, see Figure \ref{fig:non-conv}. **Explain what we see, namely example by @buur18 reproduced, showing the traceplots of chain means for some variable. The first plot is typical convergence of MICE, the second is pathological non-convergence because of a mis-specified imputation model. Each line is an imputation. In the first plot, the lines intermingle nicely and there is little trending. In the second, there is a lot of trending and the lines do not cross much. Importantly, the chain means at the last iteration (the imputed value per $m$) are very different between the two plots. The algorithm with the mis-specified model yields imputed values that are on average a magnitude two larger than those of the typically converged algorithm. This shows the importance of reaching converged states in iterative imputation algorithms.**

@@

Following the same notation, we define autocorrelation as the correlation between two subsequent $\theta$-values within the same chain [@lync07, p. 147]. In this study, we only consider $AC$ at lag 1, i.e., the correlation between the $t^{th}$ and $t+1^{th}$ iteration of the same chain.

@@

We can interpret $AC$-values as a measure of stationarity. If $AC$-values are close to zero, there is no dependence between subsequent samples within imputation chains. Negative $AC$-values indicate divergence within imputation chains. **Subsequent sampled values within each imputation chain are less alike.** Positive $AC$-values indicate recurrence. If $\theta$-values of subsequent iterations are similar, trending may occur. Negative $AC$-values show no threat to the stationarity component of convergence. On the contrary even---negative $AC$-values indicate that $\theta$-values of subsequent iterations diverge from one another, which may increase the variance of $\theta$ and speed up convergence. As convergence diagnostic, the interest is therefore in positive $AC$-values. **Maybe remove:** Moreover, the magnitude of $AC$-values may be evaluated statistically, but that is outside of this note's scope.

## In practice/Answer
**To assess whether $\widehat{R}$ and $AC$ may be appropriate non-convergence diagnostics for MICE algorithms, the following condition must hold when applied on the two algorithms plotted above: the methods indicate worse performance for the mis-specified model with pathological non-convergence (i.e., higher $\widehat{R}$- and $AC$-values than the typical performance). And additionally the methods should reflect the increasing convergence in the typical convergence situation as the number of iterations goes up.**

**In figure \ref{fig:diagnostics}A (**add panel labels**), the chain means are plotted again, now together. So are the non-convergence diagnostics. Panel B shows $\widehat{R}$ as computed by implementing @veht19 's recommendations. As required, $\widehat{R}$ indicates less signs of non-convergence as the number of iterations goes up in the typical convergence situation. The superior performance of the typical convergence over the pathological non-convergence is less prominent, even flipped for $t<4$. Panel C displays the $AC$ as computed with the R function stats::acf(). When we look at this panel, we must conclude that something weird is going on. The $AC$-values indicate equal performance (up-to $t=5$) for the typical convergence and the pathological non-convergence, while there is obvious trending in the latter. Moreover, the best convergence (as indicated by the lowest $AC$-value) is observed at $t=2$, but looking at the chain means in panel A, there should be some signs of trending up-to iteration number seven. After consulting the documentation on stats::acf(), we conclude that this implementation of $AC$ is not suitable for iterative imputation algorithms. There is a correction factor for a mathematical shortcut that works in the limit. According to... the function works when the number of iterations > 50. The default number of iterations in mice, however, is 5. Therefore, we compute $AC$ manually, see panel D. The AC values in this plot do meet the requirements (...) and will therefore be used in the simulation study.**

@@

Convergence is reached when there is no dependency between subsequent iterations of imputation chains ($AC = 0$), and chains intermingle such that the only difference between the chains is caused by the randomness induced by the algorithm ($\widehat{R} = 1$). We expect that a completely converged state will not be reached.

@@

The two challenges of convergence are stationarity and mixing. Mixing and stationarity can be inspected visually, by evaluating traceplots of scalar summaries of interest ($\theta$s; e.g., chain means and chain variances). As @buur18 describes, users can also specify a model-specific scalar summary (e.g., a regression coefficient). A user-specified scalar summary, however, is not universal to all complete data problems. Therefore, as inspired by [@mack03], we will propose a $\theta$ that summarizes the multivariate state space of the algorithm. Namely, the first eigenvalue of the variance-covariance matrix of the $M$ completed datasets.

@@

The performance of two methods ($\widehat{R}$ and $AC$) on each of these scalar summaries of interest will be evaluated with several estimands (quantities of scientific interest; $Q$s). It is assumed that when $\bar{Q}$ (the pooled result across imputations) is an unbiased (and confidence valid) estimate of $Q$, the algorithm is sufficiently converged. For each scalar $\theta$, one estimand $Q$ is defined that may be of interest in empirical research. The $\widehat{R}$ and $AC$ values with chain means as $\theta$s are evaluated against the bias in univariate mean estimates (**all variables or just for Y??**). Similarly, the performance measure for $\widehat{R}$ and $AC$ applied to chain variances is the bias in estimated standard deviations. To evaluate the performance of $\widehat{R}$ and $AC$ on the eigenvalues, we will use bias in the coefficient of determination, $R^2$. Additionally, we will evaluate the estimated regression coefficients, the coverage rate of the CI95% of the regression estimates, and the CI length (**see definitions in old Methods**). **remove the term performance measure here and add an extra paragraph to define performance measures as bias in all estimands, and coverage rate and CI length of regression coefficients.**

**Move this to Results:** As expected, conditions with a higher proportion of missingness and/or a lower number of iterations show more signs of non-convergence, as indicated by more extreme bias in the estimated $Q$s. Roughly speaking this means that MICE is indeed not converged as $t=1$, and converges gradually as $t$ increases. The point at which an additional iteration does not lead to an improvement of the estimates depends on the difficulty of the missingness problem---between XYZ and XYZ. This means that the algorithm has converged sufficiently (**under the current specifications**).

@@

Evaluation with the performance measures shows that $\widehat{R}$ and autocorrelation are conservative: they indicate signs of non-convergence in conditions where $\bar{Q}$s are unbiased and or confidence valid estimates of $Q$.

@@

Upon convergence, $\widehat{R}=1$ and $AC=0$, but in MCMC algorithms these are very unlikely (because it's a distribution, not a point to converge to). In practice, thresholds are used to diagnose non-convergence. Usually, non-convergence is diagnosed when $\widehat{R}$ > 1.2 or 1.1 or even 1.01. And a t-test is performed to assess whether $AC$ is significantly different from zero. Performance measures are the same as above: unbiased, confidence valid estimates.

**Move this to Results:** As expected, complete convergence ($\widehat{R}=1$ and $AC=0$) is not observed. But what about the thresholds for practice? $\widehat{R}$ <1.2 is not stringent enough, because conditions in which $\widehat{R}$ is smaller than 1.2 do not all yield unbiased estimates of the $Q$s. Moreover, there is a dip in the $\widehat{R}$ values, after which $\widehat{R}$ > 1.2 again. If the algorithms would be terminated at the iteration where $\widehat{R}$ < 1.2 occurs for the first time, the increase in $\widehat{R}$ values might be missed. The threshold $\widehat{R}$ < 1.1 is somewhat conservative in comparison with the performance measures. But it seems OK. $\widehat{R}$ < 1.01 is too stringent compared to the performance measures. And it is not obtained for any number of iterations (**but may be necessary in more complex complete data models???**).

@@

```
# pseudo-code of simulation 
1. simulate data 
for (missingness proportions 5%, 25%, 50%, 75% and 95%)
 for (number of simulation runs from 1 to 1000)
  2. create missingness
  for (number of iterations from 1 to 100)
   3. impute missingness
   4. compute convergence diagnostics
   5. perform analysis of scientific interest
   6. pool results across imputations
   7. compute performance measures
 8. aggregate outcomes across simulation runs 
9. combine outcomes of all missingness proportions
```
The aim of the simulation study is to evaluate the impact of inducing non-convergence by: 1) terminating the MICE algorithm at different imputation chain lengths ($t = 1, 2,..., 100$), and 2) varying the missingness proportions ($p_{miss} =.05,.25,.50,.75,.95$). The assumption underlying the different number of iterations is that the algorithm generally does reach convergence at $t=1$, because the initial values are sampled randomly from the set of observed datapoints. **As the number of iterations goes up, the imputation chains will become independent of the initial values until the point at which an extra added iteration does not lead to a more converged state.** The second set of experimental conditions under consideration--the missingness proportions--are chosen to reflect the difficulty of the missingness problem. The inherent assumption is that low missingness proportions will lead to quick algorithmic convergence, since there is a lot of information in the observed data. Higher missingness proportions then cause slower convergence. However, if the fraction of missing information is very high, there is so little information in the data that the random component in the algorithm will take the overhand and a stable but very uncertain (high variance) point will be reached.

@@

\begin{align*}
\begin{pmatrix}X_1\\
X_2\\
X_3
\end{pmatrix} \sim \mathcal{N}
\begin{bmatrix}
\begin{pmatrix}
12\\
3\\
0.5
\end{pmatrix}\!\!,
\begin{pmatrix}
4 & 4 & 1.8 & 0\\
4 & 16 & 4.8 & 0\\
1.8 & 4.8 & 9 & 0
\end{pmatrix}
\end{bmatrix}\!\!\text{.}\\[2\jot]
\end{align*} 
In this study, sampling variance is not of interest. Therefore, a single complete set may serve as comparative truth in all simulation runs [@vinknd]. A finite population of $N=1000$ is simulated using the function `mvtnorm::rmvnorm()`. Subsequently, a fourth variable is constructed to serve as dependent variable in a multiple linear regression problem. Let 
$$
Y_i = 1 + 2X_{1i} +.5X_{2i} - X_{3i} + \epsilon_i ,
$$
where $i = 1, 2,..., N$ and $\epsilon \sim \mathcal{N}(0, 100)$.

We consider four quantities of scientific interest ['conceptual estimands'; @morr19], namely... solve a multiple linear regression problem, where dependent variable $Y$ is regressed on independent variables $X_1$, $X_2$ and $X_3$ (**check notation with betas here, and add what the quantity/-ies of scientific interest is/are! Namely, the descriptive statistics of all variables (mean and standard deviation), $\beta_2$, and the percentage of variance in $Y$ explained (coefficient of determination; $R^2$)**):

@@

The complete data is 'amputed' once for each simulation repetition with function `mice::ampute()`. (**change this to reflect current simulation set-up with 5, 25, 50, 75, and 95% of cases having missing data:** The missingness is univariate, and the probability to be missing is the same for all four variables, namely 20\% (`prop = 0.8, mech = "MCAR"`). This leaves 20\% of the rows completely observed). This study only considers only an MCAR missingness mechanism. Therefore, results may not be extrapolated to other missing data problems. Proper performance of the convergence diagnostics under MCAR is necessary but not sufficient to demonstrate the appropriateness of $\widehat{R}$ and $AC$ as convergence diagnostics. Convergence diagnostics should therefore at least apply to the MCAR situation, before more complex missingness should be explored.

Missing datapoints in $y$ are imputed with the function `mice::mice()`. All MI procedures are performed with Bayesian linear regression imputation (`method = "norm"`), and five imputation chains (`m = 5`). The number of iterations varies between simulation conditions (`maxit = ` $1, 2, \dots, 100$). Each repetition of the simulation thus starts from 1 complete dataset, then splits into 5 missingness conditions ($p =.05,.25,.5,.75,.95$), each of the amputed datasets is imputed 5 times ($m = 1, 2,..., 5$) for each iteration condition ($t = 1, 2,..., 100$). From these $5\times5\times100$ imputations, we extract the $\theta$s to apply $\widehat{R}$ and $AC$ on (**i.e., chain means, chain variances, $\beta$s per imputation, and the first eigenvalue of the variance-covariance matrix per imputation**). We calculate $\widehat{R}$ by implementing Vehtari et al.'s [-@veht19] recommendations, and $AC$ as the correlation between the $t^{th}$ and the $(t+1)^{th}$ iteration. For each $\theta$ we estimate the corresponding scientific estimand $Q$.

The estimator for each $Q$ is $\bar{Q}$---the pooled aggregate of the ($\hat{Q}$)s across imputations. To calculate the $\bar{Q}$s, we combine the observed data ($y_{obs}$) and the imputed data for each imputation ($y_{imp,m}$) with the function `mice::complete()`. Descriptive statistics are computed as the average across imputations for each variable ($\mu_j$ and $\sigma_j$, where $j = Y, X_1, X_2, X_3$). Estimated regression coefficients are obtained with the function `stats::lm()` for each imputation, and then pooled conform @vink14. The coefficient of determination is estimated for each imputation, and pooled using `mice::pool.r.squared()`. 
<!-- perform multiple linear regression on each completed dataset. We obtain an estimated regression coefficient per imputation, which are pooled into a single estimate, $\bar{Q}$. We use the function `mice::pool()` to estimate u bar (**check definition!**) according to Rubin's [-@rubin87] rules, and subsequently implement finite population pooling conform @vink14. -->

The performance of $\widehat{R}$ and $AC$ is assessed by comparing $\bar{Q}$s with $Q$s. For each $Q$, we compute bias as $\bar{Q} - Q$. For $Q=\beta_j$, we also compute the empirical coverage rate (CR). CR is defined as the percentage of simulation repetitions in which the 95\% confidence interval (CI) around $\bar{Q}$ covers the true estimand $Q$. Let

$$\text{CI} = \bar{Q} \pm t_{(M-1)} \times SE_{\bar{Q}},$$

where $t_{(M-1)}$ is the quantile of a $t$-distribution with $M-1$ degrees of freedom, and $SE_{\bar{Q}}$ is the square root of the pooled variance estimate. **Remove this? ** Under-estimating the variance of $\bar{Q}$ may yield spurious inferences. Confidence interval width (CIW) is defined as the difference between the lower and upper bound of the 95\% confidence interval (CI) around $\bar{Q}$ From bias and CIW, we calculate empirical coverage rates. Coverage rate is the proportion of simulations in which $Q$ is between the bounds of the CI95\% around $\bar{Q}$.

@@

For reasons of brevity, we only discuss the convergence diagnostics for the $Q$ with the worst performance in terms of bias. For the descriptve statistics, the magnitude of the bias was the largest in $Y$ (i.e, $j = Y$ in $\mu_j$ and $\sigma_j$). For the $Q=\beta_j$, $j=2$ showed the worst perfomance (i.e, the effect of $X_1$ on $Y$.

@@

The bias in the estimates of the variable means shows little to no difference between simulation conditions. It doesn't seem to matter how many iterations you use in the mice algorithm, the estimates are unbiased. Similarly, the bias in the estimated variances is more or less stable across simulation conditions. These univariate quantities appear to be unaffected by the number of iterations.

When applied to the imputation chain means, $\widehat{R}$ indicates that the mice algorithm does not reach a converged state ($\widehat{R}$ = 1) in any of the simulation conditions. Neither is the most recent recommended threshold reached ($\widehat{R}$ < 1.01). **This is only for 75% missingness:** The conventional $\widehat{R}$ threshold of 1.2 is reached in simulation conditions $T = 3$ and $T > 6$. (With the default number of iterations (maxit = 5), this dip in $\widehat{R}$ values would be spotted, so it is no problem.) The point at which an extra iteration does not seem to improve the $\widehat{R}$ value is around $T=30$.
<!-- $\widehat{R}$ values are indistinguishable between simulation conditions when $T>30$. -->
<!-- The $\widehat{R}$ values seemed to flatten off between 20 and 30 iterations. -->

Autocorrelations indicate no sign of trending **(NOT correct anymore, fluke of the stats::acf() function!!)** within imputation chains. In most simulation conditions $AC$ is smaller than or about equal to zero. Across simulation conditions, however, the autocorrelation curve does not trend towards 0. Autocorrelation values plateau off at a value of around.1. This is a small positive autocorrelation, which would indicate some trending within chains.

The $\widehat{R}$ and $AC$ values for the imputation chain variances show equal trends to the chain means and are therefore not discussed separately. Taken together, univariate estimates seem robust across simulation conditions. There is no clear effect of the number of iterations on the bias in these estimates, while the convergence diagnostics indicate that the algorithm did not reach a completely converged state (yet).
